\hspace{1cm} Do syntactic and semantic similarity lead to interference effects?\newline Evidence from self-paced reading and event-related potentials using German

### 2023-10-14 paper/main.tex
minor edits to gen disc

Now that we have seen that the words directly preceding the critical word did not show the semantic interference effect, one might ask which brain response was elicited by the distractor itself (see Figure \ref{fig:erp_distractor})? An exploratory visual inspection of the ERPs elicited by the distractor shows several differences between conditions, but no semantic interference effect in the N400 spatio-temporal window. We believe that the differences in the ERPs elicited by the distractor in the different conditions are caused by two factors. First, the high and low syntactic interference conditions have different word orders, i.e., the distractors are embedded in different phrase structures (simple noun phrase vs.\ noun phrase including an adjective within a prepositional phrase, see Example \ref{ex:materials}). This is also apparent from the large difference between high and low syntactic interference conditions before distractor onset. Secondly, the distractors were not identical across conditions, therefore the ERPs in Figure \ref{fig:erp_distractor} are elicited by different words in the high vs.\ low semantic interference conditions. In conclusion, the distractors induced different brain responses across conditions, but these did not linger throughout the rest of the sentence (see Figure \ref{fig:erp_precrit}). %Firstly, there is a large difference prior to distractor onset (before time 0 ms) at all electrode sites. Secondly, there is a difference between 300 and 500\,ms most at anterior (left) electrode sites. We assume that both these differences are due to the different word orders in the high vs. low syntactic interference conditions, i.e., the different phrase structures that the distractors are embedded in (simple noun phrase vs. noun phrase including an adjective within a prepositional phrase, see Example \ref{ex:materials}). Thirdly, there is a late positivity (starting at approx. 500\,ms) most pronounced at centro-parietal electrodes for high semantic vs. low semantic interference conditions. This semantic interference effect in the amplitude of a late positivity could be either due to

@@

So, the comparison of the results of our SPR and ERP experiments yields that the SPR results were strongly affected by encoding interference, while the ERPs elicited by the critical verb were not. This difference between methods might be caused by the difference in presentation rate. In the self-paced reading experiment, as implied by the name, the participants had control over the presentation rate of the stimuli. The observed slowdown starting at the distractor  might have resulted from the probably subconscious intent to slow down the sampling rate of the incoming stimuli to gain more time to process them. In contrast, in the ERP experiment, we used rapid serial visual presentation (RSVP), which presents stimuli at a fixed rate that is beyond the control of the participant. Therefore, they must adapt to the set rate and process the stimuli as fast as the rate requires. Here, it is noteworthy that the presentation duration of 500\,ms per word and 400\,ms between words which was used in the present ERP experiment is slower than usual natural reading speed. However, while in SPR the slowdown due to encoding and maintaining the distractor in memory overshadowed potential retrieval effects, the procedure of the ERP experiment prevented that. In the ERP results, we observed an effect at the critical verb, which can be attributed to memory retrieval of the subject. This retrieval was only affected by the semantic cue \{± animate\} and not by the syntactic cue \{± grammatical subject\}.

>>>>

Thus, the words directly preceding the critical verb did not show the semantic interference effect. A further question that arises is: which brain response was elicited by the distractor itself (see Figure \ref{fig:erp_distractor})? An exploratory visual inspection of the ERPs elicited by the distractor shows several differences between conditions, but no semantic interference effect in the N400 spatio-temporal window. We believe that the differences in the ERPs elicited by the distractor in the different conditions are caused by two factors. First, the high and low syntactic interference conditions have different word orders, i.e., the distractors are embedded in different phrase structures (simple noun phrase vs.\ noun phrase including an adjective within a prepositional phrase; see Example \ref{ex:materials}). This is also apparent from the large difference between high and low syntactic interference conditions before distractor onset. Second, the distractors were not identical across conditions, therefore the ERPs in Figure \ref{fig:erp_distractor} are elicited by different words in the high vs.\ low semantic interference conditions. In summary, this exploratory analysis suggests that the distractors did induce different brain responses across conditions, but these could be due to factors other than encoding intererence. More importantly, these effects did not linger throughout the rest of the sentence (see Figure \ref{fig:erp_precrit}).

@@

The comparison of the results of our SPR and ERP experiments suggests that the SPR results were strongly affected by encoding interference, while the ERPs elicited by the critical verb were not. This difference between methods might be caused by the difference in presentation rate. In the self-paced reading experiment, as implied by the name, the participants had control over the presentation rate of the stimuli. The observed slowdown starting at the distractor  might have resulted from the probably subconscious intent to slow down the sampling rate of the incoming stimuli to gain more time to process them. In contrast, in the ERP experiment, we used rapid serial visual presentation (RSVP), which presents stimuli at a fixed rate that is beyond the control of the participant. Therefore, they must adapt to the set rate and process the stimuli as fast as the rate requires. Here, it is noteworthy that the presentation duration of 500\,ms per word and 400\,ms between words which was used in the present ERP experiment is slower than usual natural reading speed. However, while in SPR the slowdown due to encoding and maintaining the distractor in memory overshadowed potential retrieval effects, the procedure of the ERP experiment prevented that. In the ERP results, we observed an effect at the critical verb, which can be attributed to memory retrieval of the subject. This retrieval was only affected by the experimental manipulation.
